---
title: "Chicago Real Estate Explorer: Property Classes"
author: Charles Threlkeld
date: "11/7/2025"
fig-cap-location: margin
tbl-cap-location: margin
format:
  html:
    code-fold: true
---

## Setting the Stage

Imagine we are house hunting. We probably have a budget in mind and a general location. Let's keep our options pretty open at this point because we do not yet know what is available. Looking at listings can help, but if we give ourselves some time, other things might come onto the market or go off. If we get a handle on what is available in general, then we can have a better sense of what to expect.

Let's take a couple of examples:
1. We want to build a house from empty land in Lincoln Square.
2. We want to live in Lincoln Square or Uptown, but we do not know which is more affordable.
3. Are the types of housing the same between the two neighborhoods? That is, will we have the same range available regardless of location?
4. If we have a type of property we are interested in, then which neighborhood is a better fit?

## Getting Our Bearings

The city of Chicago is [a city of neighborhoods](https://en.wikipedia.org/wiki/List_of_neighborhoods_in_Chicago). These are geographic and cultural entities that each have their own culture and history.

For our purposes, we want to know about the different types of housing available in the neighborhoods we are interested in. The Cook County Assessor's Office provides [data](https://github.com/ccao-data/ptaxsim?tab=readme-ov-file#database-installation) that can help us answer our questions. The city also provides [geographical data](https://data.cityofchicago.org/Facilities-Geographic-Boundaries/Boundaries-Neighborhoods/bbvz-uum9) for the neighborhood boundaries.

Since we are only concerned (for now) with residential properties, we will limit our data to just residential properties, too.

So, with all that said, let's think about our first scenario: We want to build a house from empty land in Lincoln Square.

We'll go ahead and look at the data, and group it by property class:

```{r}
#| output: false
## Note on the data: my current hosting server cannot handle the size of
## ptaxsim.db, so I will be working with pre-queried data. To use this code
## with the db file, just find the commented-out queries and use them instead
## of reading from the csvs

library(bslib)
library(here)
library(knitr)
library(ptaxsim)
library(scales)
library(shiny)
library(sf)
library(tidyverse)

###############
## Constants ##
###############

## See https://prodassets.cookcountyassessoril.gov/s3fs-public/form_documents/Class_codes_definitions_12.16.24.pdf
residential_regression <- c(200, 201, 202, 203, 204, 205, 206, 207,
                            208, 209, 210, 211, 212, 234, 278, 295)
residential_nonregression <- c(200, 201, 213, 218, 219, 225, 299)
multi_family <- c(313, 314, 315, 318, 391, 396, 399)

################
## Parameters ##
################

year = 2023
location1 = "Lincoln Square"
location2 = "Uptown"

##################
## Data Sources ##
##################

## See the Assessor's Office documentation above
## This is a mapping of classes to descriptions
classes <- read_csv(here("data/class-descriptions.csv"))

## We grab the neighborhood bounds from the City Data Portal
neighborhood_bounds <- st_read("https://data.cityofchicago.org/resource/y6yq-dbs2.geojson")

sf_use_s2(FALSE)
municipality_bounds <- st_read("data/Municipality.geojson")

## And finally, our database, also supplied by the CCAO
ptaxsim_db_conn <- DBI::dbConnect(
                            RSQLite::SQLite(),
                            here("data/ptaxsim.db")
                        )

############
## Script ##
############

## residential_pins <- DBI::dbGetQuery(
##          ptaxsim_db_conn, str_glue("
## SELECT p.year, p.pin, p.class, p.av_certified, pg.longitude, pg.latitude
## FROM pin p
## INNER JOIN pin_geometry pg
## ON substr(p.pin, 1, 10) = pg.pin10
## AND p.year = pg.year
## WHERE p.year = {year}
## ")
## ) |>
##     ## Include only residental buildings
##     filter(class %in% residential_regression |  
##            class %in% residential_nonregression |
##            class %in% multi_family) |>
##     ## Map lat/lon to point
##     st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
##     ## Add in geographic data
##     st_join(neighborhood_bounds, st_within) |>
##     st_join(municipality_bounds, st_within) |>
##     st_drop_geometry() |>
##     as_tibble() |>
##     ## Apply residential multiplier
##     mutate(cash_value = av_certified * 10,
##            location = ifelse(is.na(pri_neigh), MUNICIPALITY, pri_neigh),
##            .keep = "unused") |>
##     select(class, cash_value, location) |>
##     mutate_at("class", as.factor)

residential_pins <- read_csv("data/zoning_pins.csv") |> select(class, cash_value, location)
```

```{r}
#| label: tbl-neighborhood-class-count
#| tbl-cap: "Looks like we might be out of luck looking for open land."
#| output: true
residential_pins |>
    filter(location == location1) |>
    group_by(class) |>
    summarize(count = n()) |>
    ## Add in the class English descriptions
    left_join(classes) |>
    arrange(desc(count)) |>
    kable()
```

Well, it turns out we might be out of luck looking for a parcel of land to build a house on. Looking at this list, there's not even one empty parcel of land (class 200) in the whole neighborhood! Maybe get a deal on a residential garage, but for open land, it's time to look elsewhere.

## Where to Look

Since it looks like we are not going to be building our dream house on an empty lot, let's expand our horizons a little bit. When buying a home that is already built, we are willing to live in Lincoln Square, just as before, but Uptown would be okay too. How can we tell which one is right for us?

The first question will be: which is affordable?

Let's plot a bar chart of the number of residential plots for price:

```{r}
#| label: fig-ward-comparison
#| fig-cap: "Uptown looks like it has some cheaper options than Lincoln Square."
#| warning: false

residential_pins |>
    filter(location == location1 | location == location2) |>
    ggplot(aes(cash_value, fill = location)) +
    geom_histogram(bins = 50, alpha = 0.8, position = "dodge") +
    scale_y_continuous("Number of Properties",
                       labels = label_number_auto()) +
    scale_x_continuous("Fair Cash Value",
                       labels = label_currency(prefix="$"),
                       n.breaks = 5,
                       limits = c(0, 1000000)) +
    labs(
        title = str_glue("Cash Value Distribution of Lincoln Square and Uptown in 2023"))
```

Here we can see how many properties are valued in $50,000 bins between $0 and $1M.

Just at a glance, it looks like Lincoln Square has a fairly even distribution of values, while Uptown properties tend to be largely less than $500k! The median price of a property in Lincoln Square is about $425, compared only $225 in Uptown. Quite a difference! We would save some serious cash in Uptown if we take these summary numbers at face value!

## Sorting Apples from Oranges

But we might be comparing apples and oranges. If Lincoln Square has mostly 6-unit apartment buildings and Uptown is largely single-family shotguns, we would be getting into totally different markets. So, let's look at our list from part one, but this time with both wards:

```{r}
#| label: tbl-ward-count-comparison
#| tbl-cap: "These wards are pretty different! We will want to pare down to classes we are interested in before doing any big comparisons."
#| output: true
residential_pins |>
    filter(location == location1 | location == location2) |>
    group_by(class, location) |>
    summarize(count = n()) |>
    arrange(desc(count)) |>
    pivot_wider(names_from = location,
                values_from = count,
                values_fill = 0) |>
    left_join(classes) |>
    kable()
```

Okay! So we are looking at two pretty different markets. In Uptown, the market is dominated by condominiums: over 13,000 compared to only 735 apartment buildings, the next most frequent. Lincoln Square also has many condominiums, but only about 5000. And it has many more apartment buildings than Uptown. Lincoln Square also has as many or more single family homes in nearly all classes. Uptown has many more large apartment buildings (130 vs 35), but buying a large apartment building is a relatively niche market compared to smaller, owner-occupied apartment buildings or single family residences. Good thing we did not stop with the big picture market above!

## Putting Things Back Together

Looking above, we can see that if we are in the market for a single-story residence, we might want to look at Lincoln Square, but if we want a condominium, Uptown has a much larger stock (though Lincoln Square has many, too).

But let's imagine we a 2--6 unit apartment building. Let's compare the prices between the two wards, narrowing down to only this type:

```{r}
#| label: fig-class-comparison
#| fig-cap: "Things narrowed down quite a bit, but Ward 47 still runs at a premium over Ward 40."
#| warning: false

residential_pins |>
    filter((location == location1 | location == location2) &
           class == "211") |>
    ggplot(aes(cash_value, fill = location)) +
    geom_histogram(bins = 50, alpha = 0.8, position = "dodge") +
    scale_y_continuous("Number of Properties",
                       labels = label_number_auto()) +
    scale_x_continuous("Fair Cash Value",
                       labels = label_currency(prefix="$"),
                       n.breaks = 5,
                       limits = c(0, 1000000)) +
    labs(
        title = str_glue("Cash Value Distribution of 2--6 Apartment Buildings in Lincoln Square and Uptown in 2023"))

```

This definitely tightens things up! The shapes look very similar. We are unlikely to get a significant discount in one neighborhood or the other for this type of property, but there are many more of them at nearly every price point in Lincoln Square instead of Uptown. The median value in Lincoln Square is $642k and in Uptown it is $677k, a 5% difference. But the standard deviation (a measure of how spread out the values are) is $165k and $263k respectively. So much more of the difference betweeen a property of this type in each neighbhorhood is coming from some other aspect than the neighborhood itself.

## Try It Out!

I have launched a website that allows you explore this data using whichever pair of neighborhoods (or suburbs!) you would like and whichever residential categories are of interest---even several at once! Just keep in mind that these data aren't guaranteed to be accurate to current market conditions, but should give a good sense for comparison of housing stock and relative market conditions, which are much more stable than actual prices.

You can check it out [here](https://charles-threlkeld-property-class-explorer.share.connect.posit.cloud/).
