---
title: "Chicago Real Estate Explorer: Property Classes"
author: Charles Threlkeld
date: "11/7/2025"
fig-cap-location: margin
tbl-cap-location: margin
format:
  html:
    code-fold: true
---

## Setting the Stage

In the last two explorations, I have imagined myself to be a disgruntled, home-owning taxpayer and then a fresh-faced, prospective home-buyer. However, each of these people assumes that they just take what the market gives them. On the other side of the coin are the people creating and maintaining the housing market: Developers.

A developer looks around for opportunities within the market. Like most business people, they are trying to buy low and sell high. In this case, that means taking low-value property and trying to turn it into high-value property. This will give us another lens for the housing market.

## Looking For Value

As a developer, we want to know what opportunities are available so that we can specialize. If we want to work in a location that only allows single-family homes, strip-mall specialization does not make much sense. Walking around can usually give us a pretty good idea, but let's try to be a little more objective and turn this into a quantitative question: What are the five most common zoning types in my ward and how many plots are there for each zone?

```{r}
#| output: false
library(bslib)
library(forcats)
library(here)
library(knitr)
#library(ptaxsim)
library(scales)
library(shiny)
library(sf)
library(tidyverse)

###############
## Constants ##
###############

## See https://prodassets.cookcountyassessoril.gov/s3fs-public/form_documents/Class_codes_definitions_12.16.24.pdf
residential_regression <- c(200, 201, 202, 203, 204, 205, 206, 207,
                            208, 209, 210, 211, 212, 234, 278, 295)
residential_nonregression <- c(200, 201, 213, 218, 219, 225, 299)
multi_family <- c(313, 314, 315, 318, 391, 396, 399)

################
## Parameters ##
################

year = 2023

##################
## Data Sources ##
##################

## See the Assessor's Office documentation above
## This is a mapping of classes to descriptions
classes <- read_csv(here("data/class-descriptions.csv")) |>
    mutate_at(c("class"), as.factor)

## We grab the ward bounds from the City Data Portal
ward_bounds <- st_read("https://data.cityofchicago.org/resource/k9yb-bpqx.geojson")

## Zoning code summary from
## https://secondcityzoning.org/resources/import/zoning-code-summary-district-types.csv
zoning_code_summary <- read_csv(here("data/zoning-code-summary-district-types.csv")) |>
    select(district_type_code, juan_description) |>
    mutate(zone = district_type_code,
           description = juan_description,
           .keep = "none")
    tibble()

sf_use_s2(FALSE)
## Here are the zoning details from the City ArcGIS server
## We'll do a bit of cleaning here, since we don't really need the details of special
## development zones ("PD" and "PMD") and there seem to be some typos with RM and RT
## zones
zone_bounds <- st_read("data/chicago-zoning-2025-04-17-simple.geojson") |>
    mutate(ZONE_CLASS = if_else(str_detect(ZONE_CLASS, "PD"), "PD", ZONE_CLASS)) |>
    mutate(ZONE_CLASS = if_else(str_detect(ZONE_CLASS, "PMD"), "PMD", ZONE_CLASS)) |>
    mutate(ZONE_CLASS = if_else(ZONE_CLASS == "RM4.5", "RM-4.5", ZONE_CLASS)) |>
    mutate(ZONE_CLASS = if_else(ZONE_CLASS == "RM4-.5", "RM-4.5", ZONE_CLASS)) |>
    mutate(ZONE_CLASS = if_else(ZONE_CLASS == "RM5.5", "RM-5.5", ZONE_CLASS)) |>
    mutate(ZONE_CLASS = if_else(ZONE_CLASS == "RT-4.5", "RM-4.5", ZONE_CLASS))

## And finally, our database, also supplied by the CCAO
## ptaxsim_db_conn <- DBI::dbConnect(
##                             RSQLite::SQLite(),
##                             here("data/ptaxsim.db")
                        )

############
## Script ##
############

## All Chicago tax codes start with 7. So this is a quick way to get all
## Chicago PINs and their centroids
## See https://ccao-data.github.io/ptaxsim/articles/mapping.html
## For the source of this shortcut
## residential_pins <- DBI::dbGetQuery(
##          ptaxsim_db_conn, str_glue("
## SELECT p.year, p.pin, p.class, p.av_certified, p.tax_code_num, pg.longitude, pg.latitude
## FROM pin p
## INNER JOIN pin_geometry pg
## ON substr(p.pin, 1, 10) = pg.pin10
## AND p.year = pg.year
## WHERE substr(p.tax_code_num, 1, 1) = '7'
## AND p.year = {year}
## ")
## ) |>
##     ## Map lat/lon to point
##     st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
##     ## Include only residental buildings
##     filter(class %in% residential_regression |  
##            class %in% residential_nonregression |
##            class %in% multi_family) |>
##     st_join(ward_bounds, st_within) |>  ## Add ward info
##     st_join(zone_bounds) |>  ## Add zoning info
##     st_drop_geometry() |>
##     as_tibble() |>
##     select(ZONE_CLASS, pin, class, av_certified, ward) |>
##     mutate(zone = ZONE_CLASS, 
##            cash_value = av_certified * 10,
##            .keep = "unused") |>
##    mutate_at("class", as.factor)

residential_pins <- read_csv(here("data/zone_pins.csv")) |>
    mutate_at("class", as.factor)

```

```{r}
#| label: tbl-zone-count
#| tbl-cap: "Specializing in single family homes and/or two-flats looks like a pretty good idea."
#| output: true
residential_pins |>
    filter(ward == 47) |>
    select(zone, pin, class, cash_value) |>
    group_by(zone) |>
    summarize(n = n()) |>
    arrange(desc(n)) |>
    head(n = 5) |>
    left_join(zoning_code_summary, keep = FALSE) |>
    kable()
```

Here in Ward 47, there are overwhelmingly more RS-3 zoned plots than anything else---nine times more than second-place RT-4! And looking at the rest of this abbreviated list, three of the four runner-up zones show some overlap by including single family, two-flat, or both. So, it looks like it's a good idea to specialize in building two-flats and/or single family.

But which one makes more sense for the business?

This data only shows market value, so we cannot assess building cost. Just as a thought exercise, it is likely that a single family home is less expensive than a comparably sized two-flat. There are half as many appliances, only one kitchen, and possibly fewer bathrooms. A bedroom with some simple drywall and a closet is quite a bit cheaper than the cabinetry and high-voltage electronics of a kitchen---not to mention running gas as water lines throughout, too!

However, continuing this thought experiment, one would think it could be more profitable to build a two flat. After all, if two families come along to purchase the property, their resources ought to be, on average, twice as much as a single family. But of course, this is a simplification. The people purchasing a multi-family home and the people purchasing a single family home are different target markets with different things they value and different budgets.

So, let's see how this plays out in the data. Looking at the most popular zoning classification, RS-3, we can check the price of Assessor class 207, 208, or 278 vs Assessor class 211. That is, newer two-or-more story homes with apartment buildings of 2--6 units. Since the zone only allows two-flats, I am making the assumption that the units built outside of zoning requirements are negligible in the analysis.

```{r}
#| label: fig-two-flat-vs-mult-story
#| tbl-cap: "Big single family makes a lot of money!"
#| output: true
residential_pins |>
    filter(zone == "RS-3") |>
    filter(ward == 47) |>
    filter(class %in% c(207, 208, 211, 278)) |>
    mutate(type = ifelse(class == 211, "Two-Flat", "2+ Story Single Family"), .keep = "unused") |>
    ggplot(aes(x = cash_value, y = type)) +
    geom_violin(trim = FALSE) +
    stat_summary(fun.y=mean, geom="point", shape=0, size=2, color="red") +
    stat_summary(fun.y=median, geom="point", shape=2, size=2, color="blue") +
    scale_x_continuous("Fair Cash Value",
                       labels = label_currency(prefix="$"))
```

Wow! So much for our hypothesis that a two-flat brings more money to the buyer's side of the transaction. Even if single family were more expensive to build, in this section of town, it would still be the way to go. The bulk of single family homes are valued at well over a million dollars, with some going above two million. For two-flats, one million looks more like a soft cap, with prices rarely stretching over one million. In fact, at the median and mean (blue triangles and red squares) a single family property demands a roughly $500,000 premium over a two-flat.

We'd better call our architect and make sure they only include one mail box!

But wait, maybe we are comparing apples to oranges here. For an owner-occupied two-flat, they might be limited to a single story. In the above analysis, we looked at total building size (2+ story) to try to make a comparable analysis. Instead, let's look at comparable owner-occupied living space. That is, one-story single-family properties (classes 202, 203, and 204) vs two-flats (211):

```{r}
#| label: fig-two-flat-vs-single-story
#| tbl-cap: "A single unit costs as much as two, as long as you do not have to share."
#| output: true
residential_pins |>
    filter(zone == "RS-3") |>
    filter(ward == 47) |>
    filter(class %in% c(202, 203, 204, 211)) |>
    mutate(type = ifelse(class == 211, "Two-Flat", "Single Story Single Family"), .keep = "unused") |>
    ggplot(aes(x = cash_value, y = type)) +
    geom_violin(trim = FALSE) +
    stat_summary(fun.y=mean, geom="point", shape=0, size=2, color="red") +
    stat_summary(fun.y=median, geom="point", shape=2, size=2, color="blue") +
    scale_x_continuous("Fair Cash Value",
                       labels = label_currency(prefix="$"))
```

This shows that the advantage of owning a whole second unit is basically totally offset by the premium demanded for not sharing the space. That is, in the market, people are willing to pay as much for a house to themselves as they are for *two whole units* of space that shares a floor/ceiling.

Of course, there are some mitigating factors. As an owner of a two-flat, one would be responsible for two HVAC systems, two sets of kitchen appliances, twice as many windows (though only one roof), and have to do some relational maintenance with tenants who are a bit less arms-length than a typical investment property might be. One will also need to find quality tenants that take good care of the unit and care to maintain their relationship with you, too.

There are two more things I want to check: Is this changed by ownership structure? And, are these types of units even available in this area.

To check ownership structure, we'll want an apples-to-apples comparison for a two-flat vs condominium building. Unfortunately, that's not readily available in the data. Nearly all condominium buildings are quite a bit bigger, and the units are likely smaller than we will see in a two-flat, but we can get some estimate by making the comparison.

```{r}
#| label: fig-two-flat-vs-condo
#| tbl-cap: "Condominiums are a bit cheaper than two-flats, though they are probably smaller, too."
#| output: true
residential_pins |>
    filter(zone == "RS-3") |>
    filter(ward == 47) |>
    filter(class %in% c(211, 299)) |>
    mutate(type = ifelse(class == 211, "Two Flat", "Condominium"), .keep = "unused") |>
    ggplot(aes(x = cash_value, y = type)) +
    geom_violin(trim = FALSE) +
    stat_summary(fun.y=mean, geom="point", shape=0, size=2, color="red") +
    stat_summary(fun.y=median, geom="point", shape=2, size=2, color="blue") +
    scale_x_continuous("Fair Cash Value",
                       labels = label_currency(prefix="$"))
```

Alright, that checks that box. Keep in mind that a condominium is a single unit and a two-flat is two units, so if one were to build and sell a two-flat with condominium ownership (that is, sell to two different owners with a co-ownership agreement of the shared spaces rather than sell to a single owner who would rent), there are two sales that happen. So, a single condominium averages around $250,000, and so two would average around $500,000. The price premium of a two-flat up to nearly $750,000 likely shows the premium of space, rather than ownership structure.

Finally, let's make sure that these types of units are readily available in the ward.

```{r}
#| label: tbl-class-count
#| tbl-cap: "There is currently a pretty good mix, and developers are incentivized to take it all upscale."
#| output: true
residential_pins |>
    filter(ward == 47) |>
    filter(zone == "RS-3") |>
    filter(class %in% c(202, 203, 207, 208, 211, 278, 299)) |>
    select(class, pin, class, cash_value) |>
    mutate(class = fct_collapse(class,
                                "1 Story Single Family" = c(202, 203),
                                "2+ Story Single Family" = c(207, 208, 278),
                                "Two Flat" = c(211),
                                "Condominium" = c(299))) |>
    group_by(class) |>
    summarize(n = n()) |>
    arrange(desc(n)) |>
    kable()
```

Okay, so we definitely have enough information on each type of property. There may be some mitigating effects of less availability for single family homes vs two-flats, or perhaps over supply of condominium units, and that is skewing the data. But keep in mind that everyone needs somewhere to live, and so the price premiums that we find here are reflecting the amount that people are willing to pay for one type of housing instead of the other. If there is a greater supply of one type versus this baseline, it is possible that it will lower the difference between the two and building (e.g.) new single family homes will come at the expense of tearing down two flats. And so, the two flat will become more scarce and the single family becomes more plentiful.

However, with the current market rates, as a developer, it looks quite profitable to rearrange the current two-flats into single family houses. At the median, we might have about a $500,000 budget to make the conversion. Another option is to take single-story single family homes and make them bigger. I did not show this comparison directly, but it would give us a similar budget. Single story single family buildings are relatively rare, though. We would also have to know the relative expense for conversion.

At the end of the day, though, the market is supporting more large single family homes in the neighborhood.
